Disk Heating, Galactoseismology, and the Formation of Stellar Halos

### 2017-06-18 paper/paper_v1.tex
intro edits

\documentclass[galaxies,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}
\bibliographystyle{newapa}

\usepackage{microtype}  % ALWAYS!

@@

% Math:
\newcommand{\msun}{\mathrm{M}_\odot}
\newcommand{\kms}{\mathrm{km}~\mathrm{s}^{-1}}

@@

Emerging in the 1990's, the catalogues that inspired this recent acceleration not only mapped global structures in our Galaxy, but also revealed the ubiquity of substructure within it.
These revelations added an unforeseen richness to interpretations, and encouraged the development of new dynamical tools for studying ongoing interactions and formation histories:
\project{Hipparcos} \citep{esa97} found moving groups of stars in the velocity distribution in the Solar Neighborhood \citep{dehnen98}, some corresponding (as expected) to destroyed stars clusters, while others (unexpectedly) could plausibly be explained as signatures of resonances with the Galactic bar \citep{dehnen00};
the Sloan Digital Sky Survey \citep[hereafter, SDSS ---][]{york00,stoughton02,abazajian03} found many streams of Main Sequence Turnoff (MSTO) stars in our stellar halo, thought to be from long-dead satellite galaxies \citep{newberg02,belokurov06} as a stunning confirmation that our Galaxy had indeed formed hierarchically \citep[e.g.][]{bullock01,bullock05};
and M giant stars selected from the Two Micron All Sky Survey \citep[hereafter, 2MASS ---][]{nikolaev00} traced the debris from the Sagittarius dwarf galaxy entirely around the Sky \citep{majewski03} to offer a window on the history of its disruption \citep{law10}.

\begin{figure}[t]
\label{fig:ting}
\centering
%\includegraphics[width=3 cm]{logo-mdpi}
\caption{\label{fig:ting}
Summary of spatial distributions of M giants}
\end{figure}

Figure \ref{fig:ting}, reproduced from \cite{li17}, shows the spatial distribution of M giant stars associated with three such substructures that this paper concentrates on. Note that this map was made using estimates from photometry  for the distances to these stars from the Sun.
The expected distance uncertainties using this method are of order 20\% \citep{sheffield14}, so any fine features discussed below will not be apparent.
These substructures were found in SDSS and 2MASS as deviations from the expected global structure of the disk or inner stellar halo around the anti-center and at a range of low to moderate Galactic latitudes --- we will hereafter refer to them collectively as the ``low-latitude structures''.
Note that the same region on the sky has been shown to be richly structured on even smaller  scales\citep{slater14,martin14,deason14} but we do not discuss this sub-substructure here.
\begin{description}
\item{\it The Monoceros Ring or Giant Anticenter Stellar Structure(hereafter Mon/GASS)} is a distinct ring of stars beyond the expected edge of the Galactic disk \citep[at $\sim$ 5 kpc beyound the Sun][]{robin92},  spanning the longitude range ???, with Galactic latitudes ??? and distances from the Sun $\sim$?? kpc. It was originally found as excess of main sequence stars in SDSS \citep{yanny03,ibata03} and M giant stars in 2MASS \citep{rochapinto03}. Follow-up spectra of the M giant stars indicate that they follow a clear trend in mean velocity with Galactic longitude with small dispersion \citep{crane04}.
\item{\it The Triangulum-Andromeda Clouds, I and II,} were first discovered as a single diffuse over-density of M giant stars, covering the area ????, overlapping the Mon/GASS structure on the sky, but at distances of approximately 15-20 kpc from the Sun \citep{rochapinto04}.
Spectra showed that the M giants exhibited a coherent velocity structure with small dispersion  \citep{rochapinto04} and  deep photometry in the region allowed the clear identification of MSTO stars and an estimate of a surface brightness $\Sigma >$ 32 mag/arcsec$^2$ \cite{majewski04}.
Subsequent photometric work indicated the presence of {\it two} distinct MSTO \citep{martic07}.
\item{\it A13} is another tenuous association of stars discovered by applying a group finding algorithm \citep{sharma09} to the catalogue of all 2MASS M giants \citep{sharma10}.
This group sits in the north Glalactic hemisphere in the area ???? on the sky and at approximate distances of ???.
\end{description}
There are many models for the formation of these substructures.
Mon/GASS has attributed to the accretion of a satellite \citep[on a retrograde orbit][]{penarrubia05, michel-dansac11}, a natural extension of the Galactic warp \citep{momany04,momany06} or disturbances to the Galactic disk \citep{kazantzidis08,younger08,purcell11,gomez16}.
For TriAnd I/II and A13, their extreme position at (R,Z)$\sim$(kpc,kpc) seems to exclude a disturbed disk as a possible origin in these cases, and a satellite on a retrograde orbit has again been shown to provide a plausible explanation \citep{sheffield14}.

A more convincing, coherent picture of the nature of these low-latitude structures is just beginning to emerge.
This paper reviews the contributions our own group has made to formulating this picture, which include surveys of the low-latitude structures in spectra \citep{sheffield14,ting17}, stellar populations \citep{pricewhelan15,sheffield17} and abundance patterns \citep{bergemann17}, as well as numerical simulations \citep{sheffield14,laporte17a,laporte17b}.
We summarize this observational and theoretical work in Section 2 and 3 respectively, adding in the context of contemporary work from other groups, as well as the larger context of possible connections across the Galactic disk.
Armed with this understanding of the nature of these substructures, we go on in Section 4 to discuss prospects for mapping such structures more generally around our own and other Galaxies.
We end in Section 5 by outlining the motivation for making such maps, asking what they might be telling us about bigger questions: galaxy formation scenarios and dark matter distributions.

@@

We have recently obtained high-resolution and signal-to-noise spectra of 15 stars in TriAnd1 and A13 overdensities. 14 stars were observed with the HIRES-S spectrograph at the Keck-1 telescope \cite{vogt1994} and 1 star was observed using the the UVES spectrograph at the VLT (Program ID: 097.B-0770A). The spectral resolving power R of the HIRES spectra is 36\,000 and the UVES data have R $\sim$ 47\,000. All Keck spectra cover the full optical region, from $4800$ to $8770$ $\Angstrom$, although the exact wavelength coverage varies as several slightly different instrument configurations were tried in an effort to get all the key lines into a single exposure. The signal-to-noise (SNR)/\Angstrom of the HIRES spectra near 5200 \Angstrom in the continuum at the center of the echelle order exceeds 200. For the UVES spectrum, the SNR of 50 near 5500 \Angstrom was achieved. The fundamental parameters and chemical abundances of stars were determined using 2MASS and APASS photometry and the high-resolution spectra. All 15 stars are M-type giants with effective temperatures Teff ~ 3800 K and surface gravities log(g) ~ 1 dex. The $T_{\rm eff}$ estimates were derived using the Infrared Flux Method (Casagrande et al. 2010, 2014). Chemical abundances were measured for 6 chemical elements, including O, Na, Mg, Ca, Fe, Eu using LTE and the standard MARCS stellar model atmospheres (Gustafsson et al. 2008). We have also performed detailed analysis of the effects of non-local thermodynamic equilibrium (non-LTE) for the elements, where detailed calculations are available \cite{bergemann2011, bergemann2012, bergemann2016}, however, the NLTE corrections for the chosen lines are minor and do not change our conclusions. In what follows, we use our LTE results, because all studies of chemical abundances in dSph systems and most studies in the Milky Way to date have employed LTE with 1D hydrostatic models

We find that stars in the TriAnd1 and in A13 overdensities have extemely similar chemical abundances, with the abundance dispersion $\leq 0.06$ dex for most chemical elements. The TriAnd1 and A13 stars also have a very narrow metallicity spread,  [Fe/H] $= -0.92 \pm 0.06$ dex. The abundances of all measured alpha-elements\footnote{The only exception is [Ca/Fe], which is solar with the dispersion of 0.06 dex; however, the NLTE abundance correction for the \ion{Ca}{i} lines for M giants is 0.1 dex that will bring the results in agreement with the MW disk studies that are based on FGK dwarfs, for which the NLTE correction is much smaller \cite{Merle2011}}. are uniformly enhanced at the level of 0.43 $\sim$ 0.06 dex for [O/Fe], 0.41 $\sim$ 0.02$ dex for [Mg/Fe], $0.35 \pm 0.06$ dex for [Na/Fe], and 0.15 $\sim$ 0.03 dex for [Eu/Fe] (Figure 1).However, the At this level, the TriAnd1 and A13 stars are consistent with the abundances of the in-situ formed Milky Way thick disk stars \cite{fuhrmann2002,bergemann2014,bensby2014}. The abundance ratios are too high for the chemical abundance patterns observed in the stars of the Galactic satellites (Bonifacio et al. 2010, Shetrone et al. 2001, 2003, Tolstoy et al. 2009, de Boer et al. 2014), which are known to have low, typically solar ( [O/Fe], [Mg/Fe]) or even sub-solar ([Na/Fe]), ratios at [Fe/H] $\sim -1$. For example, Fornax, a dSph galaxy most close to TriAnd1 and A13 in metallicity, has [Na/Fe] $\sim -0.6$ dex and [Mg/Fe] $\sim 0$ at [Fe/H] $=-1$ (Letarte et al. 2010, Kirby 2010, Lemasle et al. 2014) \footnote{Note that to avoid systematic differences between abundance measurements obtained from spectra taken with different instruments, we have chosen here to compare with the data taken with the same instrument; here UVES and FLAMES at the VLT, ESO, or with HIRES at Keck.}. Thus it is unlikely that the TriAnd1 and A13 stars originate from a disrupted dSph galaxy.

Our results for TriAnd and A13 do not confirm the conclusions by \cite{chou2010b}. Although there could be several astrophysical reasons for the differences, this could be caused by the differences in the observed datasets and spectroscopic modelling techniques. Our observed data cover the full optical range from the near-UV to the near-IR, whereas \cite{chou2010b} have analysed only the small wavelength region in the near-IR, limited to 150 \Angstrom from 7440 to 7590 \Angstrom. Because of this limitation, they could include only 11 \ion{Fe}{I} and 2 \ion{Ti}{I} lines in the determination of gravities, metallicities, and abundances. Our analysis, through a much wider wavelength coverage and high SNR attained for the observed spectra, allowed us to include 70 lines of \ion{Fe}{I} and \ion{Fe}{II} \cite{bergemann2012}, as well as 10 lines of \ion{Ti}{2}. \cite{bergemann2011} showed that \ion{Ti}{1} should be not be used in abundance studies because it is very sensitive to NLTE effects. We have also performed test computations, using a reduced linelist, to explore the sensitivity of abundance diagnostic to the line selection and wavelength regimes. We have found that using the linelist from \cite{chou2010a}, the metallicities are over-estimated by $\sim 0.25$ dex and [Ti/Fe] are under-estimated by $0.4$ dex, compared to the results using the complete linelist. This suggests that the choice of the linelist and diagnostic spectral band (full optical or near-IR) could be a possible reason for the difference between us and \cite{chou2010}.
%{\it Summary: the abundance patterns  of the low-latitude structures are similar to those of the thick disk of our Galaxy.}

@@

%=====================================
% References, variant A: internal bibliography
%=====================================
\bibliography{refs_1.bib,refs_2.bib,refs_3.bib}

>>>>

\documentclass[galaxies,article,submit,moreauthors,pdftex,10pt,a4paper]{mdpi}

\usepackage{microtype}  % ALWAYS!
\usepackage{amsmath}
\usepackage{amsrefs}
\usepackage{aas_macros}

@@

% Math:
\newcommand{\kpc}{\mathrm{kpc}}
\newcommand{\msun}{\mathrm{M}_\odot}
\newcommand{\kms}{\mathrm{km}~\mathrm{s}^{-1}}

\newcommand{\ion}[1]{#1}

% APW notes to self:
% - Hercules-Aquila, false-positive?

@@

Emerging in the 1990's, the catalogues that inspired this recent acceleration not only mapped global structures in our Galaxy, but also revealed the ubiquity of substructure within it.
These revelations added an unforeseen richness to interpretations, and encouraged the development of new dynamical tools for studying ongoing interactions and formation histories:
\survey{Hipparcos} \citep{esa97} found moving groups of stars in the velocity distribution in the Solar Neighborhood \citep{dehnen98}, some corresponding (as expected) to destroyed stars clusters, while others (unexpectedly) could plausibly be explained as signatures of resonances with the Galactic bar \citep{dehnen00};
the Sloan Digital Sky Survey \citep[hereafter, SDSS ---][]{york00,stoughton02,abazajian03} found many streams of Main Sequence Turnoff (MSTO) stars in our stellar halo, thought to be from long-dead satellite galaxies and dissolved globular clusters \citep{newberg02,belokurov06} as a stunning confirmation that our Galaxy had indeed formed hierarchically \citep[e.g.][]{bullock01,bullock05};
and M giant stars selected from the Two Micron All Sky Survey \citep[hereafter, 2MASS ---][]{nikolaev00} traced the debris from the Sagittarius dwarf galaxy entirely around the Sky \citep{majewski03} to offer a window on the history of its disruption \citep{law10}.
From these and many other studies that use large survey catalogues, it is now clear that the Milky Way is full of kinematic substructure.
% APW COMMENT: this section makes it sound like the surveys discovered these things "for free," but really it was the authors you cite who found these things. I would rearrange to say, e.g., "Using data from the \survey{Hipparcos} mission \citep{esa97}, moving groups of stars were found in the velocity distribution in the Solar Neighborhood \citep{dehnen98}...etc"

\begin{figure}[t]
\label{fig:ting}
\centering
%\includegraphics[width=3 cm]{logo-mdpi}
\caption{\label{fig:ting}
Summary of spatial distributions of M giants
The distance measurements are from photometry alone and have expected absolute uncertainties of $\approx$20\% \citep{sheffield14}.
}
\end{figure}
% APW COMMENT: Maybe Ting would be willing to share her data, and I can re-make
% the figure? You know how I am about visualizations... :)

Figure~\ref{fig:ting}, reproduced from \cite{li17}, shows the spatial distribution of M giant stars associated with three such substructures that this paper concentrates on: the so-called Monoceros Ring, Triangulum-Andromeda clouds, and A13.
%Any fine features discussed below will not be apparent.
Each of these substructures were originally identified using SDSS and 2MASS photometry as over-densities in stellar number counts relative to the expected global structure of the disk or inner stellar halo.
Unlike stellar streams, these substructures are present at a range of low to moderate Galactic latitudes --- we will hereafter refer to them collectively as the ``low-latitude structures.''
Note that the same region on the sky has been shown to be richly structured on even smaller scales\citep{slater14,martin14,deason14} but we do not discuss these sub-substructures here.
\begin{description}
\item{\it The Monoceros Ring or Giant Anticenter Stellar Structure} (hereafter Mon/GASS) is an arc- or ring-like feature of stars beyond the expected edge of the Galactic disk \citep[assumed to be $\approx 5~\kpc$ beyond the Sun in cylindrical radius][]{robin92}.
The feature spans a large area: Galactic longitudes $\approx 120^\circ \lesssim l \lesssim 240^\circ$, latitudes $-30^\circ \lesssim b \lesssim +40^\circ$, and Heliocentric distances $5\lesssim d_\odot \lesssim 10~\kpc$ (\citealt{Morganson:2016}).
Mon/GASS was originally found as an excess of main sequence stars in SDSS \citep{yanny03,ibata03} and M giant stars in 2MASS \citep{rochapinto03}.
Follow-up spectra of the M giant stars indicate that they follow a clear trend in mean velocity with Galactic longitude with small dispersion \citep{crane04}.
\item{\it The Triangulum-Andromeda Clouds I and II} (hereafter TriAnd 1 and 2) were first discovered as a single diffuse over-density of M giant stars covering the area $\approx 100^\circ \lesssim l \lesssim 150^\circ$, $\approx 20^\circ \lesssim b \lesssim 40^\circ$ overlapping the Mon/GASS structure on the sky but at larger Heliocentric distances of approximately $15$--$30~\kpc$ \citep{rochapinto04}.
Follow-up spectroscopy showed that the M giants exhibit a coherent velocity sequence with a small dispersion \citep{rochapinto04} and later deep photometry in the region revealed associated MSTO stars and an estimate of the surface brightness of the feature $\Sigma > 32~{\rm mag}~{\rm arcsec}^{-2}$ \cite{majewski04}.
Subsequent photometric follow-up then revealed the presence of {\it two} distinct main sequences \citep{martin07}.
\item{\it A13} is another tenuous association of stars discovered by applying a group finding algorithm \citep{sharma09} to the catalogue of all 2MASS M giants \citep{sharma10}.
This group sits in the north Galactic hemisphere in the area $\approx 125^\circ \lesssim l \lesssim 210^\circ$, $\approx 20^\circ \lesssim b \lesssim 40^\circ$ and at approximate distances of $\approx 12$--$20~\kpc$.
\end{description}
We note that (1) these structures likely continue towards lower Galactic
latitudes ($b < 20^\circ$) but are incomplete due to Galactic extinction near
the midplane, and (2) each structure in reality has a longitude-dependent
distance dependence that is ignored in the quoted distance ranges above.
Many models have been proposed to explain the formation and existence of these
substructures.
Mon/GASS has been attributed to the accretion of a satellite \citep[on a retrograde orbit][]{penarrubia05, michel-dansac11}, a natural extension of the Galactic warp \citep{momany04,momany06}, or disturbances to the Galactic disk \citep{kazantzidis08,younger08,purcell11,xu15,gomez16}.
For TriAnd I/II and A13, their extreme positions at $(R,Z) \approx (20, -8)~\kpc$, $(R,Z) \approx (25, -10)~\kpc$, and $(R,Z) \approx (20, 8)~\kpc$ (across a large range in Galactocentric azimuth, $\phi$) seems to exclude a disturbed disk as a possible origin in these cases, and a satellite on a retrograde orbit has again been shown to provide a plausible explanation \citep{sheffield14}.
% APW COMMENT: I found it strange that the historical context doesn't include
% our recent work - maybe we should state that this was the status before our
% group started working on follow-up, or something? Otherwise, GASS section
% should get a reference to upcoming Sheffield+(in prep.) paper on RR Lyrae,
% TriAnd section should get references to Sheffield+2014(?) and
% Price-Whelan+2015, A13 paper should get reference to Li+2017

A more convincing, coherent picture of the nature of these low-latitude structures is just beginning to emerge.
This paper reviews the contributions our own group has made to formulating this picture, which include spectroscopic surveys of the low-latitude structures to study metallicity and velocities \citep{sheffield14,ting17}, stellar populations \citep{pricewhelan15,sheffield17}, and abundance patterns \citep{bergemann17}, as well as numerical simulations \citep{sheffield14,laporte17a,laporte17b}.
We summarize this observational and theoretical work in Section 2 and 3 respectively, adding in the context of contemporary work from other groups, as well as the larger context of possible connections across the Galactic disk.
Armed with this understanding of the nature of these substructures, we go on in Section 4 to discuss prospects for mapping such structures more generally around our own and other Galaxies.
We end in Section 5 by outlining the motivation for making such maps, asking what they might be telling us about bigger questions: galaxy formation scenarios and dark matter distributions.

@@

We have recently obtained high-resolution and signal-to-noise spectra of 15 stars in TriAnd1 and A13 overdensities. 14 stars were observed with the HIRES-S spectrograph at the Keck-1 telescope \cite{vogt1994} and 1 star was observed using the the UVES spectrograph at the VLT (Program ID: 097.B-0770A). The spectral resolving power R of the HIRES spectra is 36\,000 and the UVES data have R $\sim$ 47\,000. All Keck spectra cover the full optical region, from $4800$ to $8770$ $\AA$, although the exact wavelength coverage varies as several slightly different instrument configurations were tried in an effort to get all the key lines into a single exposure. The signal-to-noise (SNR)/\AA of the HIRES spectra near 5200 \AA in the continuum at the center of the echelle order exceeds 200. For the UVES spectrum, the SNR of 50 near 5500 \AA was achieved. The fundamental parameters and chemical abundances of stars were determined using 2MASS and APASS photometry and the high-resolution spectra. All 15 stars are M-type giants with effective temperatures Teff ~ 3800 K and surface gravities log(g) ~ 1 dex. The $T_{\rm eff}$ estimates were derived using the Infrared Flux Method (Casagrande et al. 2010, 2014). Chemical abundances were measured for 6 chemical elements, including O, Na, Mg, Ca, Fe, Eu using LTE and the standard MARCS stellar model atmospheres (Gustafsson et al. 2008). We have also performed detailed analysis of the effects of non-local thermodynamic equilibrium (non-LTE) for the elements, where detailed calculations are available \cite{bergemann2011, bergemann2012, bergemann2016}, however, the NLTE corrections for the chosen lines are minor and do not change our conclusions. In what follows, we use our LTE results, because all studies of chemical abundances in dSph systems and most studies in the Milky Way to date have employed LTE with 1D hydrostatic models

We find that stars in the TriAnd1 and in A13 overdensities have extemely similar chemical abundances, with the abundance dispersion $\leq 0.06$ dex for most chemical elements. The TriAnd1 and A13 stars also have a very narrow metallicity spread,  [Fe/H] $= -0.92 \pm 0.06$ dex. The abundances of all measured alpha-elements\footnote{The only exception is [Ca/Fe], which is solar with the dispersion of 0.06 dex; however, the NLTE abundance correction for the \ion{Ca}{i} lines for M giants is 0.1 dex that will bring the results in agreement with the MW disk studies that are based on FGK dwarfs, for which the NLTE correction is much smaller \cite{Merle2011}}. are uniformly enhanced at the level of 0.43 $\sim$ 0.06 dex for [O/Fe], 0.41 $\sim$ 0.02 dex for [Mg/Fe], $0.35 \pm 0.06$ dex for [Na/Fe], and 0.15 $\sim$ 0.03 dex for [Eu/Fe] (Figure 1).However, the At this level, the TriAnd1 and A13 stars are consistent with the abundances of the in-situ formed Milky Way thick disk stars \cite{fuhrmann2002,bergemann2014,bensby2014}. The abundance ratios are too high for the chemical abundance patterns observed in the stars of the Galactic satellites (Bonifacio et al. 2010, Shetrone et al. 2001, 2003, Tolstoy et al. 2009, de Boer et al. 2014), which are known to have low, typically solar ( [O/Fe], [Mg/Fe]) or even sub-solar ([Na/Fe]), ratios at [Fe/H] $\sim -1$. For example, Fornax, a dSph galaxy most close to TriAnd1 and A13 in metallicity, has [Na/Fe] $\sim -0.6$ dex and [Mg/Fe] $\sim 0$ at [Fe/H] $=-1$ (Letarte et al. 2010, Kirby 2010, Lemasle et al. 2014) \footnote{Note that to avoid systematic differences between abundance measurements obtained from spectra taken with different instruments, we have chosen here to compare with the data taken with the same instrument; here UVES and FLAMES at the VLT, ESO, or with HIRES at Keck.}. Thus it is unlikely that the TriAnd1 and A13 stars originate from a disrupted dSph galaxy.

Our results for TriAnd and A13 do not confirm the conclusions by \cite{chou2010b}. Although there could be several astrophysical reasons for the differences, this could be caused by the differences in the observed datasets and spectroscopic modelling techniques. Our observed data cover the full optical range from the near-UV to the near-IR, whereas \cite{chou2010b} have analysed only the small wavelength region in the near-IR, limited to 150 \AA from 7440 to 7590 \AA. Because of this limitation, they could include only 11 \ion{Fe}{I} and 2 \ion{Ti}{I} lines in the determination of gravities, metallicities, and abundances. Our analysis, through a much wider wavelength coverage and high SNR attained for the observed spectra, allowed us to include 70 lines of \ion{Fe}{I} and \ion{Fe}{II} \cite{bergemann2012}, as well as 10 lines of \ion{Ti}{2}. \cite{bergemann2011} showed that \ion{Ti}{1} should be not be used in abundance studies because it is very sensitive to NLTE effects. We have also performed test computations, using a reduced linelist, to explore the sensitivity of abundance diagnostic to the line selection and wavelength regimes. We have found that using the linelist from \cite{chou2010a}, the metallicities are over-estimated by $\sim 0.25$ dex and [Ti/Fe] are under-estimated by $0.4$ dex, compared to the results using the complete linelist. This suggests that the choice of the linelist and diagnostic spectral band (full optical or near-IR) could be a possible reason for the difference between us and \cite{chou2010}.
%{\it Summary: the abundance patterns  of the low-latitude structures are similar to those of the thick disk of our Galaxy.}

@@

%=====================================
% References, variant A: internal bibliography
%=====================================
% \bibliographystyle{chicago2}
\bibliography{refs_1.bib,refs_2.bib,refs_3.bib,refs_apw}
